\sys: Automatic Disease Diagnosis Using Large Language Models and Answer Set Programming

### 2025-09-12 paper/draft.tex
Update intro

\documentclass[10pt,leqno]{amsart}
\usepackage{graphicx}
\baselineskip=16pt
\usepackage{indentfirst,csquotes}

\topmargin= .5cm
\textheight= 20cm
\textwidth= 32cc
\baselineskip=16pt

\evensidemargin= .9cm
\oddsidemargin= .9cm

@@

The development of automatic disease prediction systems 
is crucial for enhancing healthcare and supporting medical 
professionals in timely and accurate diagnosis. 
Symbolic artificial intelligence (AI), particularly ASP, 
has been explored in this context due to its ability 
to support structured and explainable reasoning. 
However, the construction of expressive knowledge bases 
remains a significant barrier to practical adoption.

This paper introduces \sys, a framework 
that combines LLMs with the 
formal reasoning capabilities of ASP.
\sys extracts and encodes medical knowledge from text into ASP code,
enabling interpretable and accurate diagnosis 
of diseases based on patient symptoms.

\section{Background}

@@

        \node[draw, dotted, fit=(grounder) (solver), inner sep=2mm, label=above:{Solving}] {};
\end{tikzpicture}
    \end{center}
		\caption{Solving a problem using ASP.}
    \label{fig:asp-solving}
\end{figure}

@@

More recent work explores the synergy between LLMs and ASP 
for enhanced diagnostic reasoning. 
These hybrid approaches aim to combine the language understanding of LLMs 
with the precision and explainability of symbolic reasoning. 
A YAML-based interface has been proposed to guide LLMs 
in extracting structured facts from natural language, 
which are then processed by ASP for logical inference \cite{alviano2024llm2asp}. 
Other work uses ASP to validate and trace misleading or inconsistent outputs 
from LLMs in medical contexts \cite{Nguyen2025}. 
A specialized lightweight LLM has also been trained to generate valid ASP code, 
improving the automation of rule-based medical reasoning \cite{coppolillo2024llasp}.
Collectively, these systems assist in building reliable and clear reasoning pipelines 
for disease diagnosis by combining language understanding with rule-based validation.
\section{Methodology}

@@

\section{Results}

@@

\section{Conclusion}

>>>>

\documentclass[11pt,leqno]{amsart}
\usepackage{graphicx}
\baselineskip=16pt
\usepackage{indentfirst,csquotes}
\usepackage[margin=1in]{geometry}

% \topmargin= .5cm
% \textheight= 20cm
% \textwidth= 32cc
% \baselineskip=16pt

% \evensidemargin= .9cm
% \oddsidemargin= .9cm

@@

Accurate and timely diagnosis is critical for effective treatment and patient outcomes, 
but building automated systems that can deliver this reliably remains challenging. 
Symbolic AI---and ASP in particular---offers transparent reasoning, 
but adoption has been hindered by the high cost of 
authoring and maintaining rich, domain-specific knowledge bases at scale. 
Meanwhile, LLMs show strong performance on medical text 
but often produce unreliable or unstructured outputs, limiting their use in 
high-stakes clinical settings.

\sys addresses this gap by coupling LLMs with ASP to automatically construct 
explainable diagnostic programs from medical literature. 
\sys uses prompting to translate disease-specific literature into ASP rules, 
augments these rules with patient symptoms and indicators, 
and invokes an ASP solver to produce candidate diagnoses together with explicit logical justifications.

Concretely, this work:
\begin{itemize}
    \item Builds a framework that turns medical text into an executable ASP knowledge base.
    \item Uses prompts to produce modular rules (including alternatives and symptom links) for better explainability and partial inference.
    \item Adds solver-based explanations to show clear symptom $\rightarrow$ rule $\rightarrow$ diagnosis chains.
    \item Demonstrates accurate results on real patient data while keeping the reasoning transparent.
\end{itemize}

On publicly available patient records, \sys achieves 
95–-100\% accuracy across selected diseases, while producing transparent explanations for each diagnosis.

This paper begins with background on LLMs and ASP (\cref{sec:background}), 
then presents the \sys methodology (\cref{sec:methodology}), 
evaluates the framework on patient data (\cref{sec:results}),
and concludes with limitations and future directions (\cref{sec:conclusion}).

\section{Background}\label{sec:background}

@@

        \node[draw, dotted, fit=(grounder) (solver), inner sep=2mm, label=above:{Solving}] {};
\end{tikzpicture}
    \end{center}
		\caption{\textbf{Solving a problem with ASP}. A problem is first modeled as a program, then transformed into ground rules by a grounder. A solver computes the stable models, which are interpreted as solutions.}
    \label{fig:asp-solving}
\end{figure}

@@

More recent work explores the synergy between LLMs and ASP 
for enhanced diagnostic reasoning. 
These hybrid approaches aim to combine the language understanding of LLMs 
with the precision and explainability of symbolic reasoning. 
A YAML-based interface has been proposed to guide LLMs 
in extracting structured facts from natural language, 
which are then processed by ASP for logical inference \cite{alviano2024llm2asp}. 
Other work uses ASP to validate and trace misleading or inconsistent outputs 
from LLMs in medical contexts \cite{Nguyen2025}. 
A specialized lightweight LLM has also been trained to generate valid ASP code, 
improving the automation of rule-based medical reasoning \cite{coppolillo2024llasp}.
Collectively, these systems assist in building reliable and clear reasoning pipelines 
for disease diagnosis by combining language understanding with rule-based validation.

\section{Methodology} \label{sec:methodology}

@@

\section{Results}\label{sec:results}

@@

\section{Conclusion}\label{sec:conclusion}
